Social Visual Analytics: DevOps for Data Scientists

### 2015-01-27 introduction.tex
working

%% \todo{Points we want to make."Coordination is done by meetings" "how do you trace an automated alert back to an EDA environment ?"}

% What's the area
Consider the emerging role of a data science team within an
organization today \cite{Keim:2008:VAS}.
%
% How do people do it today?
Individual data scientists and statisticians usually work on loosely
related problems, and must devise effective ways to share their findings
and move results from exploratory data analysis to automated diagnostics
and reports that are deployed for wider consumption.
They generally rely on a patchwork of resources shared informally,
or at best, through systems that attack individual tasks or functions
in the data analysis workcycle. % might cite information visualization reference model or pirolli/card

% Why does this not work?
There are two problems with the current practice. First, there are
gaps in the workflow: Exploratory Data Analysis (EDA) is done with one
set of tools, and automated reports and deployments with another. This
makes transferring results between EDA and production (in both
directions) slow, expensive, and error-prone. Coordination is done
through meetings and may take months to carry out.

Second, EDA environments often assume a single-developer perspective,
while data scientist teams could greatly benefit from convenient searching
and sharing of scripts, data sets, data feeds, experiments, annotations,
and automated analyses and recommendations. These capabilities have had
tremendous impact on information retrieval in general but are well
beyond what existing visual analytics environments provide, or what
is found in today's repositories and version control
systems.\cite{Koren:2009:FIT}

In addition, consider trends in interactive visualization software
in the open source community.  Projects led by Jeff Heer
include Prefuse, Flare, Protovis, d3, and Vega, in that order.
All but the first of these libraries take advantage of web technologies,
and all but the second rely on open standards. Hadley Wickham is the
author of ggplot, ggplot2, and ggvis. ggplot and ggplot2 generate
static graphics, while ggvis provides convenient interactive visualizations
in R, using the R package Shiny and the Vega abstract visualization grammar,
opening a separate web server for each call. These projects show
a trend toward visual analytics frameworks and systems that are
increasingly rich, interactive, composable, and accessible through the web.

% What are we going to do about this?

%% \todo{Define the requirements for the system}
%% We begin by noting the main requirements for a data analysis
%% environment for collaboration and transfer of results into production.

%% Main requirement is to satisfy these two sets of users: R coders and
%% managers. They work differently and want different things. What should
%% the infrastructure look like so they both coexist in peace? If this is
%% such a fundamental problem, why have other tools not addressed it?
%% Discuss Shiny (maybe in the disc. section).

Today's state-of-the-art environments for data science teams are unquestionably
powerful, but also fragmented. Instead of creating a new solution that promotes
even further fragmentation, we attempt to identify and remedy the shortcomings
in the current practice of data science. The following requirements summarize
the main pain points:

@@

\item Technology transfer. In our experience, development
of new analyses and visualizations is done by \emph{hackers}
and \emph{scripters} (using Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}) who are, by definition,
at least somewhat familiar with programming and software development.
On the other hand, \emph{application
users} ultimately derive value from the data by means of the tools
developed by hackers and scripters. Traditionally, the connection
between these two worlds is made by IT staff. In an environment where
business needs can change every day, manual IT support does not scale:
we need to create software infrastructure to connect the two worlds.
%% The two classes of users
%% have quite different perspectives but we
%% infrastructure for continuous deployment of visualizations.

\item Coexistence. In the current data science ecosystem, the
isolation of exploratory visualization and data analysis
environments is a hinderance to wider adoption of modern techniques
on both sides. We specifically see the richness of interactive
visualization tools as still being isolated from the richness
of statistical programming environments.
%(The path backward from production to analysis can be 
%equally important: when problems discovered in
%production require expert help, we would like experts to be
%able to immediately apply the full range of analytic tools without
%extracting data to move it back to the EDA environment.)
There is an opportunity to provide a framework so that developments
on either side are more easily employed by the other.
In addition, the integrated results should be more easily accessible
to the rest of the organization.

\item Discoverability. From our experience in working with data
analyst teams, we find that repeated work (``How can I connect to
database X?'', ``How do you get clean data from column Y from feed
Z?'') arises from lack of communication about previously solved subproblems.
We wish to provide transparency about the work of team members in the organization.

@@

Technologies which fail to address these requirements
are holding back the practice and impact of interactive visual
analysis. In this paper, we discuss the design and deployment of
RCloud, a prototype environment that supports collaborative data
analysis, visualization and web deployment satisfying all requirements
above. We will report on the design decisions, tradeoffs and
limitations, and compare the concepts in RCloud to related proposals.

%% Our investigation focuses on the R language \cite{RCoreTeam:2013:R}
%% for statistical computing and graphics. R is one of the most popular
%% systems for data analysis and is well-supported by
%% a complex ecosystem of compatible packages, software
%% tools and other resources.
%
%% Of importance to us, there are existing
%% packages for high performance computing in the cloud,
%% and for generating web graphics. % no, there aren't. we wrote those!

RCloud uses the R language~\cite{RCoreTeam:2013:R} as the core set
of statistical analysis tools.  The facilities for \emph{sharing, viewing
and embedding} RCloud notebooks address the {\bf technology transfer} requrement.
Because RCloud embraces technologies that are part of the computing
environment (specifically, inter-process communication via HTTP and
web services as remote procedure calls, as explained
in Section~\ref{sec:highlevelarchitecture}), it is straightforward to
integrate RCloud with satellite efforts in corporate environments.

{\bf Coexistence} between analysis and visualization is supported in
its human interface architecture, described in~\ref{sec:humaninterface}. 
As oulined in Section~\ref{sec:notebooks}, RCloud notebooks are sequences
of R snippets interspersed with Markdown for easy generation of reports.
We describe how RCloud enables interactive visualization widgets in
this setting (Sections~\ref{sec:Rinbrowser} and~\ref{sec:interactivenotebooks}).
Indexing, publishing and sharing of notebooks is also the main way
Rcloud supports {\bf discoverability}.

The requirement of continuous technology transfer, though, creates challenges
about the \emph{stability} of the many moving pieces: if notebooks are
updated and deployed automatically, how can the system handle
changing requirements and public interfaces? We discuss our choice of
transparent, integrated version control for RCloud notebooks in 
Section~\ref{sec:deployment}.

Section~\ref{sec:casestudy} describes a case study illustrating some of these concepts.

In Section~\ref{sec:limitations}, we report on experience toward meeting
these requirements in RCloud, and limitations encountered.

%% Dashboards are a solution to requirement 1, using solutions of
%% requirement 2.
%% Requirement 2: \emph{how is accessibility supposed to happen?}

%% Version control as a primary concept in the environment is a
%% consequence of requirement 1. There
%% needs to be a way to control the visibility of prototypes as they are
%% being developed and tested.

%% Tight integration between R process and web browser is a solution to
%% requirement 2

%% Current isolation notwithstanding, we find there is demand for interactive
%% visualization tools. We do not believe that moving the \

%% The main challenge here is how to integrate these
%% technologies so data scientists and statisticians can target
%% them, \emph{and} business-facing users can use them.
%% Scalable technologies like crossfilter exist. Want to support those in the context of an R environment. Need to support full 2-way communication between computing in the cloud and interaction on the web.
%% \item Other requirements e.g. Kandel's user study. Not our contribution but things we must address, mentioned in related work section. Heer/Agrawala design considerations for collaborative visual analytics.

% What is the scope of our solution?

% We want our solution to play well with the rest of the ecosystem

>>>>

Consider the role of a data science team within an organization today.
Analysts work in loose or tight collaboration with colleagues.
Project teams vary in size, from just a few people to dozens or
more, even within one project's duration. Tasks are broad, including problem
identification, data wrangling, visualization, modeling and analysis,
interpreting, explaining and advocating the results, and often helping
clients to realize the benefits of the work in ongoing processes.
There are many details to these tasks.
For example, data wrangling can involve finding data, figuring out its
syntax and semantics, assessing data quality, performing normalization and
data quality remediation, and making the data available for other tasks
to follow.

Visual analytics in practice depends greatly on communication and collaboration.
At almost every step, detailed knowledge about data, code and tasks
is shared by collaborators. 
Further, data scientists are increasingly asked to work more closely
with business people who are domain specialists and less technically oriented.
Thus, data scientists and developers are being asked to become
very broad, and integrate work across the spectrum. (Hiring ``full stack
developers'' to perform integration seems to be a trend, but does not achieve
the potential benefits of specialization and division of labor in larger
teams, and is not a practice to be encouraged.)
This pushes the boundaries for communication and collaboration even further.
Unfortunately, processes and technology that support this work are fragmented.
Knowledge is shared through informal conversations, email online messages,
meetings, phone calls, within code itself, in wikis, in project documents and
by other means. Results of experiments are often shared first by copying static
output, later, by holding meetings that lead to moving or even rewriting
entire programs.

This disjointed process inhibits the use of tools to bridge
the many gaps and to help people find resources.
For example, when code is shared, we still not may know where to find examples
that show how to use it. stephen{Why did I shift to first person?}
When we find a useful data source, we don't know what packages or
other data sources usually go along with it, or what problems with the data
others have already discovered and corrected, or where to obtain online feeds
to get updates of the data.

The current situation affects how well data science teams work. It limits
the ability to share work and to rapidly deploy and refine results.
We have seen this in our lab environment.
Other authors report similar findings.

Some symptoms of the current situation are:
\begin{itemize}
\item It is hard to find data, metadata, and previously hard-won knowledge about it.
\item Most coordination is done through meetings, whose content is not related to other artifacts and may not even be stored.
\item Production deployment requires porting code to a different environment or even completely rewriting it, thwarting continuous release.
\item Many tools adopt a standalone or single developer perspective,
not suited to collaboration and web deployment. 
\item Analysts find insufficient support for tracing events or issues from production back to the EDA process
\end{itemize}

In view of the opportunity to improve this situation,
we created a software environment that supports the end to end
visual analytics process for individuals and teams.
This environment knits together some familiar tools, and it
provides new features to find data and code; create experimental workbooks;
run experiments; annotate and deploy experiments as end-user websites or
as reusable, callable services; and to share, search and recommend these artifacts. The artifacts are stored in a version
control system that provides a common workspace, as well as needed control
and isolation between stable and experimental versions of code and other
resources.

Over the past two years, we prototyped this environment and deployed it
in a production community of data scientists, business analysts and other colleagues.
The platform today has over 200 active accounts, and several dozen people have forked
and customized code in the system.

A key point is that, for the most part, the improved capabilities are
available by default, without much explicit intervention on the part of
data scientists and other customers. Visual analytics experiments are
conveniently shared and turned into production websites, without moving
or porting code. All the published artifacts in the system can be
searched immediately. Recommendation is as easy as clicking a star
on a useful workbook.

The high level architecture of the prototype system
is shown in figure~\ref{fig:high-level-architecture}.
We chose R as the foundation for our prototype because it is already
the dominant statistical computing language in our lab.
R also has many useful packages for data analysis and visualization,
and the core system and its packages are open source that can be modified to
support research.

There is much previous work that inspired or influenced ours. Some
software prototypes that we learned from include:

{\bf Vistrails} as an important prototype of a scientific workflow
and provenance management system. The prototype helped the community
to understand the value of capturing aspects of the processes that
surround data analysis experiments and tools, including detailed
history, collaboration, and deployment.

{\bf iPython notebooks (Jupyter)} are the closest work we are aware of.
Of particular note,

{\bf RStudio, RShiny} and other packages based on R for creating and
publishing web content such as knitR, Rpubs, rCharts, and gg2v.

{\bf bl.ocks, jsfiddle, plotly} and other web services for sharing code
and demos.

{ManyEyes} is the best-known foundational work showing the potential for
integrating data visualization with social media.

Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
data from lab experiments and sharing it within research teams \cite{Rubacha:ELN:2011}. In a sense, we hope to adapt and extend this concept to the work of
visual analytics teams.

From a more principled or theoretical perspective, visual analytics goals
or objectives for emerging systems, described in previous work, closely match
the goals of our prototype.  Adopting the terminology of X Y and Z, we noted
the following.

@@

\item Technology transfer.
In most organizations, development of analyses and visualizations 
s done by \emph{hackers} and \emph{scripters} (using Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
gain the benefits of the tools developed by hackers and scripters.
Generally, the connection between these two worlds is made by IT staff,
who port or even rewrite code so it can run as a stable production service.
In an environment where business needs can change rapidly, this
process does not scale. Our objective is to connect the two worlds
of EDA and production.

\item Coexistence. In the current data science ecosystem, the
isolation of exploratory visualization and data analysis
environments hinders wider adoption of modern techniques.
The richness of interactive visualization tools is still somewhat
separated from the power of statistical programming environments.
There is an opportunity to provide a framework so that developments
on each side are more easily adopted by the other, and made available
in production services.

\item Discoverability. A chronic complaint of data analysis teams is
repeated work (``How can I connect to database X?'', ``How do you get
clean data from column Y from feed Z?'')  This work arises from lack of
communication about previously solved subproblems.  Our goal is encourage
and support transparency of work between team members.

@@

\stephen{We could now tell them what we're going to tell them here, but I'm
not sure that's a good approach - unless we feel the reader is having
trouble following the path we are laying out, it seems to disrupt the
flow of the discussion.}
